Introducing ICBe: An Event Extraction Dataset from Narratives about International Crises

### 2024-02-16 replication_paper/psrm_draft/ICBe_manuscript.Rmd
conditional accept edits to fig 1 and 3 and sec 1.1 and 5

How can scholars abstract and measure discrete events about a historical episode in international relations? We employ a metaphor of chess, a game that despite its complexity can nonetheless be recorded in a standardized and structured manner. Like chess, international interactions involve a finite set of players (e.g. polities, rebel groups, IGOs) expressing their behaviors (e.g. thinking, saying, and doing) by moving pieces (e.g. military platforms, civilian personnel, diplomats) from and onto identifiable locations (geo-coded coordinates). These moves occur over a marked time period (start and end of a historical episode) and gambits occur in recordable sequences (actions and reactions) to produce observable, if still disputable, outcomes (e.g. victory, defeat, stalemate, peace). Our knowledge of this historical episode, including the actors involved as well as their preferences, behaviors, and beliefs are only indirectly observed from historical records that most often take the form of unstructured natural language text.

Much like the recording of chess evolved from natural language descriptive text notation to the modern figurine algebra notion, international relations scholars have recently sought to produce a structured account of historical events by combining the unstructured corpus of historical records with informative priors about international relations. The resulting structured account of information about a historical episode can be combined with that of other events to produce a systematic account of political events from which we can garner novel insight, assuming the underlying data have high coverage, precision, and recall. The easiest way to convey the desired produced of this task is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) in natural language sentences alongside a mapping to discrete machine-readable abstractive events. From the structured data, scholars can identify similarities and differences across events concerning important concepts like when particular foreign policy actions deter versus inflame [@jervisCooperationSecurityDilemma1978; @glaserCausesConsequencesArms2000], when third parties mediate in interstate disputes [@haffarEmergentPeacemakersCataloguing2002; @quinnPowerPlayMediation2006], and how actors try to communicate resolve [@tragerDiplomacyWarPeace2016; @luptonReexaminingReputationResolve2018]. Identifying patterns of international interactions is not just an inherently interesting enterprise; it is a necessary precondition to important efforts to predict where policymakers should turn their attention to improve global welfare [@wardLearningSteppingFuture2013; @begerReassessingRoleTheory2021].

@@

![Cuban Missile Crisis (1962) - ICB Narrative vs. ICBe Events](case_study_cuban_precision.png){#fig-cubanarrative width="7in" height="8in"}

@@

![Dots represent individual ICB narrative sentences, as embedded by the Paraphrase-MPNET-base-v2 large language model and flattened into two dimensions with UMAP. Text labels reflect individual leaves of the ICBe ontology, and colors represent intermediate branches of the ontology. Label placement is the median of all of the sentences that the tag was applied to by the coders. The dendrogram shows hierarchical clustering of the tags. If ICBe precision is high, then the sentences that tags were applied to ought to say similar things and the intended shape of the ontology ought to be visually recognizable.](p_semantic_embeddings_dendro.png){#fig-umap width="7in" height="7in"}

@@

We investigated event abstraction from narratives describing key historical episodes in international relations. We synthesized a prior belief about the latent unobserved phenomena that drive these events in international relations and proposed a mapping to observable concepts that enter into the observed historical record. We designed an ontology with high coverage over those concepts and developed a training procedure and technical stack for human coding of historical texts. Multiple validity checks find the resulting codings have high internal validity (e.g. intercoder agreement) and external validity (i.e. matching source material in both micro-details at the sentence level and macro-details spanning full historical episodes). Further, these codings perform much better in terms of recall, precision, coverage, and overall coherence in capturing these historical episodes than existing event systems used in international relations.

We release several open-source products along with supporting code and documentation to further advance the study of international relations, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts).[^13] Second, we provide a coder-level disaggregation with multiple codings of each sentence by experts and undergrads that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here at crisisevents.org.

[^13]: Using ICBe data, @gannonOneIfLand2022 finds that cross-domain crises are shorter and less violent than same-domain crises.

>>>>

Our knowledge of any historical episode, including the participants and their preferences, behaviors, and beliefs are only indirectly observed from historical records that most often take the form of unstructured natural language text. Despite its complexity, all international interactions fundamentally involve a finite set of actors expressing their interests through at least theoretically observable behaviors. So how can scholars abstract and measure discrete events about a historical episode in international relations? Scholars have recently sought to produce a structured account of world events by combining the unstructured corpus of historical records with informative priors about international relations. The result can produce novel insight from a systematic account of political events. The easiest way to convey the desired product is with an example. Figure 1 shows a narrative account of the Cuban Missile Crisis (1962) in natural language sentences alongside a mapping to discrete machine-readable abstractive events. From this, scholars can identify similarities and differences across events like what foreign policy actions deter versus inflame [@jervisCooperationSecurityDilemma1978; @glaserCausesConsequencesArms2000], when third parties mediate [@haffarEmergentPeacemakersCataloguing2002; @quinnPowerPlayMediation2006], and how actors communicate resolve [@tragerDiplomacyWarPeace2016; @luptonReexaminingReputationResolve2018]. Identifying patterns of international interactions is not just an inherently interesting enterprise; it is a necessary precondition to important efforts to predict where policymakers should turn their attention to improve global welfare [@wardLearningSteppingFuture2013; @begerReassessingRoleTheory2021].

@@

![Comparison of a natural language and machine-readable abstractive account of the Cuban Missile Crisis (1962). The text on the left is a summary of the event from the ICB Crisis Narrative. The mapping on the right shows the corresponding ICBe coding](case_study_cuban_precision.png){#fig-cubanarrative width="7in" height="8in"}

@@

![Computational evaluation of the precision of ICBe event codings. The plot on the left is a map of the semantic meaning of every sentence in the corpus (black points) as assigned by a large language model (Paraphrase-MPNET-base-v2) and projected down into two dimensions (UMAP). Overlaid are the average semantic locations of each label assigned by ICBe coders (colored labels). The labels with similar meaning are assigned to sentences with similar semantic meaning, creating an observable structure and pattern we would not observe with low quality coding where tag location would instead appear random. The plot on the right shows a hierarchical dendrogram clustering labels into groups by their average semantic location with more similar labels being more closely connected on the tree. The clustering by color indicates it closely mirrors the intended ICBe ontology, suggesting high precision in the coding.](p_semantic_embeddings_dendro.png){#fig-umap width="7in" height="7in"}

@@

The scope and complexity of international politics should not discourage the identification of trends, patterns, and regularities. In undertaking event abstraction from narratives about key historical episodes in international relations, this paper has proposed a mapping between unstructured historical records and a structured ontology of these events with high coverage of concepts of interest. Multiple validity checks find the resulting codings have high internal validity (e.g. intercoder agreement) and external validity (i.e. matching source material in both micro-details at the sentence level and macro-details spanning full historical episodes). Further, these codings perform much better in terms of recall, precision, coverage, and overall coherence in capturing these historical episodes than existing event systems used in international relations.

These data, along with the open-source code, documentation, and companion website provide several substantive and methodological contributions to the discipline. Substantively, these data are appropriate for statistical analysis of hard questions in the study of crises like interactions between means of warfare and the preconditions for conflict escalation [@gannonOneIfLand2022]. Methodologically, our mapping from codings to source text at the sentence level provide a new resource for natural language processing with access to coder-level disaggregation that furthers the study of uncertainty in the interpretation of international events and in the quantitative coding of historical events. Finally, we provide a companion website (crisisevents.org) that incorporates detailed visualizations of all the data introduced here as a new resource for the study of international crises in a scalable, yet detailed, manner.

### 2024-02-16 replication_paper/psrm_draft/ICBe_manuscript.tex
conditional accept edits to fig 1 and 3 and sec 1.1 and 5

How can scholars abstract and measure discrete events about a historical
episode in international relations? We employ a metaphor of chess, a
game that despite its complexity can nonetheless be recorded in a
standardized and structured manner. Like chess, international
interactions involve a finite set of players (e.g.~polities, rebel
groups, IGOs) expressing their behaviors (e.g.~thinking, saying, and
doing) by moving pieces (e.g.~military platforms, civilian personnel,
diplomats) from and onto identifiable locations (geo-coded coordinates).
These moves occur over a marked time period (start and end of a
historical episode) and gambits occur in recordable sequences (actions
and reactions) to produce observable, if still disputable, outcomes
(e.g.~victory, defeat, stalemate, peace). Our knowledge of this
historical episode, including the actors involved as well as their
preferences, behaviors, and beliefs are only indirectly observed from
historical records that most often take the form of unstructured natural
language text.

Much like the recording of chess evolved from natural language
descriptive text notation to the modern figurine algebra notion,
international relations scholars have recently sought to produce a
structured account of historical events by combining the unstructured
corpus of historical records with informative priors about international
relations. The resulting structured account of information about a
historical episode can be combined with that of other events to produce
a systematic account of political events from which we can garner novel
insight, assuming the underlying data have high coverage, precision, and
recall. The easiest way to convey the desired produced of this task is
with an example. Figure 1 shows a narrative account of the Cuban Missile
Crisis (1962) in natural language sentences alongside a mapping to
discrete machine-readable abstractive events. From the structured data,
scholars can identify similarities and differences across events
concerning important concepts like when particular foreign policy
actions deter versus inflame (Jervis 1978; Glaser 2000), when third
parties mediate in interstate disputes (Haffar 2002; Quinn et al. 2006),
and how actors try to communicate resolve (Trager 2016; Lupton 2018).
Identifying patterns of international interactions is not just an
inherently interesting enterprise; it is a necessary precondition to
important efforts to predict where policymakers should turn their
attention to improve global welfare (Ward et al. 2013; Beger, Morgan,
and Ward 2021).

\begin{figure}
\hypertarget{fig-cubanarrative}{%
\centering
\includegraphics[width=7in,height=8in]{case_study_cuban_precision.png}
\caption{Cuban Missile Crisis (1962) - ICB Narrative vs.~ICBe
Events}\label{fig-cubanarrative}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig-umap}{%
\centering
\includegraphics[width=7in,height=7in]{p_semantic_embeddings_dendro.png}
\caption{Dots represent individual ICB narrative sentences, as embedded
by the Paraphrase-MPNET-base-v2 large language model and flattened into
two dimensions with UMAP. Text labels reflect individual leaves of the
ICBe ontology, and colors represent intermediate branches of the
ontology. Label placement is the median of all of the sentences that the
tag was applied to by the coders. The dendrogram shows hierarchical
clustering of the tags. If ICBe precision is high, then the sentences
that tags were applied to ought to say similar things and the intended
shape of the ontology ought to be visually
recognizable.}\label{fig-umap}
}
\end{figure}

@@

We investigated event abstraction from narratives describing key
historical episodes in international relations. We synthesized a prior
belief about the latent unobserved phenomena that drive these events in
international relations and proposed a mapping to observable concepts
that enter into the observed historical record. We designed an ontology
with high coverage over those concepts and developed a training
procedure and technical stack for human coding of historical texts.
Multiple validity checks find the resulting codings have high internal
validity (e.g.~intercoder agreement) and external validity
(i.e.~matching source material in both micro-details at the sentence
level and macro-details spanning full historical episodes). Further,
these codings perform much better in terms of recall, precision,
coverage, and overall coherence in capturing these historical episodes
than existing event systems used in international relations.

We release several open-source products along with supporting code and
documentation to further advance the study of international relations,
event extraction, and natural language processing. The first is the
International Crisis Behavior Events (ICBe) dataset, an event-level
aggregation of what took place during the crises identified by the ICB
project. These data are appropriate for statistical analysis of hard
questions about the sequencing of events (e.g.~escalation and
de-escalation of conflicts).\footnote{Using ICBe data, Gannon (2022)
  finds that cross-domain crises are shorter and less violent than
  same-domain crises.} Second, we provide a coder-level disaggregation
with multiple codings of each sentence by experts and undergrads that
allows for the introduction of uncertainty and human interpretation of
events. Further, we release a direct mapping from the codings to the
source text at the sentence level as a new resource for natural language
processing. Finally, we provide a companion website that incorporates
detailed visualizations of all of the data introduced here at
crisisevents.org.

>>>>

Our knowledge of any historical episode, including the participants and
their preferences, behaviors, and beliefs are only indirectly observed
from historical records that most often take the form of unstructured
natural language text. Despite its complexity, all international
interactions fundamentally involve a finite set of actors expressing
their interests through at least theoretically observable behaviors. So
how can scholars abstract and measure discrete events about a historical
episode in international relations? Scholars have recently sought to
produce a structured account of world events by combining the
unstructured corpus of historical records with informative priors about
international relations. The result can produce novel insight from a
systematic account of political events. The easiest way to convey the
desired product is with an example. Figure 1 shows a narrative account
of the Cuban Missile Crisis (1962) in natural language sentences
alongside a mapping to discrete machine-readable abstractive events.
From this, scholars can identify similarities and differences across
events like what foreign policy actions deter versus inflame (Jervis
1978; Glaser 2000), when third parties mediate (Haffar 2002; Quinn et
al. 2006), and how actors communicate resolve (Trager 2016; Lupton
2018). Identifying patterns of international interactions is not just an
inherently interesting enterprise; it is a necessary precondition to
important efforts to predict where policymakers should turn their
attention to improve global welfare (Ward et al. 2013; Beger, Morgan,
and Ward 2021).

\begin{figure}
\hypertarget{fig-cubanarrative}{%
\centering
\includegraphics[width=7in,height=8in]{case_study_cuban_precision.png}
\caption{Comparison of a natural language and machine-readable
abstractive account of the Cuban Missile Crisis (1962). The text on the
left is a summary of the event from the ICB Crisis Narrative. The
mapping on the right shows the corresponding ICBe
coding}\label{fig-cubanarrative}
}
\end{figure}

@@

\begin{figure}
\hypertarget{fig-umap}{%
\centering
\includegraphics[width=7in,height=7in]{p_semantic_embeddings_dendro.png}
\caption{Computational evaluation of the precision of ICBe event
codings. The plot on the left is a map of the semantic meaning of every
sentence in the corpus (black points) as assigned by a large language
model (Paraphrase-MPNET-base-v2) and projected down into two dimensions
(UMAP). Overlaid are the average semantic locations of each label
assigned by ICBe coders (colored labels). The labels with similar
meaning are assigned to sentences with similar semantic meaning,
creating an observable structure and pattern we would not observe with
low quality coding where tag location would instead appear random. The
plot on the right shows a hierarchical dendrogram clustering labels into
groups by their average semantic location with more similar labels being
more closely connected on the tree. The clustering by color indicates it
closely mirrors the intended ICBe ontology, suggesting high precision in
the coding.}\label{fig-umap}
}
\end{figure}

@@

The scope and complexity of international politics should not discourage
the identification of trends, patterns, and regularities. In undertaking
event abstraction from narratives about key historical episodes in
international relations, this paper has proposed a mapping between
unstructured historical records and a structured ontology of these
events with high coverage of concepts of interest. Multiple validity
checks find the resulting codings have high internal validity
(e.g.~intercoder agreement) and external validity (i.e.~matching source
material in both micro-details at the sentence level and macro-details
spanning full historical episodes). Further, these codings perform much
better in terms of recall, precision, coverage, and overall coherence in
capturing these historical episodes than existing event systems used in
international relations.

These data, along with the open-source code, documentation, and
companion website provide several substantive and methodological
contributions to the discipline. Substantively, these data are
appropriate for statistical analysis of hard questions in the study of
crises like interactions between means of warfare and the preconditions
for conflict escalation (Gannon 2022). Methodologically, our mapping
from codings to source text at the sentence level provide a new resource
for natural language processing with access to coder-level
disaggregation that furthers the study of uncertainty in the
interpretation of international events and in the quantitative coding of
historical events. Finally, we provide a companion website
(crisisevents.org) that incorporates detailed visualizations of all the
data introduced here as a new resource for the study of international
crises in a scalable, yet detailed, manner.
